\documentclass[final, twoside]{IEEEtran} % regular paper
\setlength\columnwidth{0.5\textwidth} % this must be set in one-column mode
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{float}
\usepackage{amssymb, amsmath, amsthm, bm}
\usepackage{enumitem}
\newcommand{\Mod}[1]{\ (\mathrm{mod}\ #1)}
\makeatletter
\newcommand{\tpmod}[1]{{\@displayfalse\pmod{#1}}}
\makeatother
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\iu}{{i\mkern1mu}}
\usepackage{graphicx, color}
\graphicspath{{figures-pdf/}}
\usepackage{cite}
\usepackage{url}
\usepackage{diagbox}
\usepackage[aboveskip=1pt]{subcaption}
\usepackage{pbox}
\usepackage{enumitem}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}

\usepackage{algpseudocode}
\usepackage{multirow, bigstrut}
\usepackage{tabularx}
\usepackage{arydshln}
\usepackage{empheq}
\usepackage{threeparttable}

%%%% TIKZ PACKAGES FOR FIGURES %%%%
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.geometric}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{fit}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.pathreplacing}

%%%% COLOR PACKAGE FOR HIGHLIGHTING %%%%
\usepackage{xcolor}
\usepackage[normalem]{ulem}
\newcommand{\bluet}[1]{\textcolor{blue}{#1}}

\newlength\OneImW
\setlength\OneImW{0.38\columnwidth}

\newlength\BigOneImW
\setlength\BigOneImW{0.85\columnwidth}

\newlength\twofigwidth
\setlength\twofigwidth{0.46\columnwidth}

\newlength\ThreeImW
\setlength\ThreeImW{0.31\columnwidth}

\newlength\sfigwidth
\setlength\sfigwidth{0.3\columnwidth}

\newlength\vfigskip
\setlength\vfigskip{4em}

\newcommand\mymatrix[1]{\bm{\mathrm{#1}}}
\newcommand{\sign}[1]{\mathrm{sgn}{#1}} 

\hyphenation{op-tical net-works semi-conduc-tor}

\newlength\figsep
\setlength\figsep{1.5em}

\usepackage{stfloats} 

\newtheorem{Proposition}{Proposition}
\newtheorem{Definition}{Definition}
\newtheorem{Corollary}{Corollary}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\DeclareMathOperator*{\lcm}{lcm}
%%% Vector and matrix operators
\newcommand{\x}[0]{{\bf x}}
\newcommand{\vct}[1]{\bm{#1}}
\newcommand{\mtx}[1]{\bm{#1}}

%%% Constants, vectors, and matrices with names
\newcommand{\atom}{\vct{\phi}}
\newcommand{\Fee}{\mtx{\Phi}}

\usepackage[bookmarks=false]{hyperref}
\hypersetup{
 linktocpage=true, pdfborderstyle={/S/S/W 1}, hyperindex=true, bookmarks=true, bookmarksopen=true, bookmarksnumbered=true
}

\begin{document}

\title{ML-DSA Digital Signatures in Resource-Constrained MQTT Environments}

\author{ Jiahao Xiang, Lang Li  and Jingya Feng

	\thanks{This research is supported by the Open Fund of Hunan Engineering Research Center for Cyberspace Security Technology and Application at Hengyang Normal University (2025HSKFJJ031), "the 14th Five-Year Plan" Key Disciplines and Application-oriented Special Disciplines of Hunan Province (Xiangjiaotong [2022] 351), the Science and Technology Innovation Program of Hunan Province (2016TP1020). \textit{(Corresponding
			author: Lang Li.)}}

	\thanks{The authors are with the Hunan Provincial Key Laboratory of Intelligent Information Processing and Application, the Hunan Engineering Research Center of Cyberspace Security Technology and Applications, and the College of Computer Science and Technology, Hengyang Normal University, Hengyang 421002, China (e-mail: jiahaoxiang2000@gmail.com; lilang911@126.com; fengjyk@126.com).}

}

% The paper headers
\markboth{IEEE Transactions}{Xiang \MakeLowercase{et al.}}

\maketitle

\begin{abstract}
	The emergence of large-scale quantum computers makes migration of Internet of Things (IoT) systems to post-quantum cryptographic standards increasingly necessary. This paper studies the deployment of ML-DSA (Module-Lattice-Based Digital Signature Algorithm) in MQTT-based IoT systems on ARM Cortex-M4 microcontrollers from three connected perspectives: implementation optimization, MQTT-oriented signature integration, and system-level evaluation. An optimized ML-DSA implementation is developed by targeting the dominant arithmetic kernels on constrained devices, and a broker-compatible method is used to embed ML-DSA signatures into MQTT publish-subscribe communication. Experimental results indicate that ML-DSA still imposes substantial overhead relative to ECDSA, with signature generation latencies of \textbf{663--1,136\,ms} and computational overhead of \textbf{70.6--121.0$\times$} across parameter sets. End-to-end evaluation further quantifies message expansion, latency, throughput, and memory cost, providing practical guidance for ML-DSA deployment in resource-constrained MQTT environments.
\end{abstract}

\begin{IEEEkeywords}
	Post-Quantum Cryptography, ML-DSA, MQTT Protocol, IoT Security, Resource-Constrained Devices
\end{IEEEkeywords}

\section{Introduction}

\IEEEPARstart{T}{he} emergence of quantum computing undermines widely deployed public-key cryptographic primitives and makes migration to post-quantum standards a practical requirement rather than a long-term theoretical concern~\cite{Khalid2019,Dinu2025,Egbuagha2025}. This transition is especially important for Internet of Things (IoT) systems, where large numbers of distributed devices must operate for long periods under strict constraints on computation, memory, energy, and communication bandwidth.

\IEEEpubidadjcol

Among post-quantum primitives, ML-DSA (Module-Lattice-Based Digital Signature Algorithm) has been standardized by the National Institute of Standards and Technology (NIST) in FIPS 204~\cite{NIST-FIPS-204} as the primary digital signature scheme for post-quantum deployment. ML-DSA is particularly relevant in settings that require public verifiability, non-repudiation, auditability, and compatibility with certificate-based trust models. Recent work has therefore focused on implementation portability across embedded software, vectorized ARM processors, and hardware accelerators~\cite{Barrett2023,Wei2025,Norga2025}. These properties, however, come with substantial implementation cost. ML-DSA signatures span 2,420--4,627 bytes across standardized parameter sets, which is roughly 30--70$\times$ larger than 64-byte ECDSA signatures~\cite{NIST-FIPS-204}. In addition, the underlying lattice-based computations impose significantly higher processing overhead on low-power devices~\cite{Ghosh2019,Benchmark2022,pqm4-benchmarks}.

These costs become more critical in MQTT-based IoT environments. MQTT is widely adopted because its publish-subscribe architecture and lightweight packet format suit resource-constrained sensing and control applications. The same design choices, however, also make the protocol sensitive to cryptographic expansion. When multi-kilobyte signatures are attached to short telemetry payloads, message size, transmission delay, and signing latency can dominate end-to-end performance. As a result, the feasibility of deploying standardized post-quantum signatures in practical MQTT workflows cannot be inferred from algorithm specifications alone and must be evaluated at the system level.

Existing studies have examined post-quantum signatures from several complementary perspectives. Embedded benchmarking and optimization studies have reported the cost of Dilithium and ML-DSA on ARM Cortex-M platforms and related microcontrollers, including Cortex-M3, Cortex-M4, Cortex-M7, and 8-bit AVR targets~\cite{Banegas2021,Benchmark2022,pqm4-benchmarks,Barrett2023}. Recent deployment-oriented work has also highlighted migration engineering and implementation-security issues for ML-DSA, including fault-attack exposure and countermeasure requirements~\cite{Li2024,Dinu2025}. At the protocol level, MQTT-focused studies have shown that post-quantum authentication can substantially increase latency and payload size~\cite{Samandari2023,Rampazzo2023}, protocol adaptations have been explored for MQTT-SN and CoAP~\cite{BlancoRomero2024}, and KEM-based alternatives have been proposed for highly constrained MQTT nodes~\cite{Kim2025}. Broader reviews of protocol migration likewise identify messaging systems as an area where deployment evidence remains limited~\cite{Egbuagha2025}. Nevertheless, a focused evaluation of standardized ML-DSA parameter sets within a complete MQTT publish-subscribe workflow on Cortex-M4-class microcontrollers remains limited. In particular, the relationship among cryptographic cost, message expansion, and protocol-level delay requires clearer empirical characterization.

This work addresses that gap through an empirical study of ML-DSA integration in MQTT-based IoT systems on ARM Cortex-M4 microcontrollers. The study is organized as a connected chain from implementation optimization to protocol integration and then to system-level evaluation. The main contributions are as follows:

\begin{enumerate}
\item \textbf{Implementation-Level Optimization}: An optimized ML-DSA implementation is developed for ARM Cortex-M4 microcontrollers by targeting the dominant arithmetic kernels that determine execution cost in resource-constrained environments.

\item \textbf{MQTT-Oriented Signature Integration}: A broker-compatible integration method is designed to embed ML-DSA signatures into MQTT publish-subscribe communication while preserving practical deployability on constrained devices.

\item \textbf{System-Level Evaluation and Deployment Analysis}: Cycle-accurate and end-to-end experiments quantify computational overhead, message expansion, latency, throughput, and memory cost across all three standardized ML-DSA parameter sets relative to an ECDSA P-256 baseline, providing guidance for constrained IoT deployment.
\end{enumerate}

The remainder of this paper is organized as follows: Section~\ref{sec:related} presents technical background. Section~\ref{sec:architecture} describes the optimized implementation and MQTT-oriented integration method. Section~\ref{sec:evaluation} presents experimental evaluation. Section~\ref{sec:conclusion} concludes with implications for deployment.

\section{Background}\label{sec:related}

\subsection{ML-DSA and Deployment-Relevant Properties}\label{sec:complexity}

ML-DSA is NIST's standardized post-quantum digital signature scheme in FIPS 204~\cite{NIST-FIPS-204} and is derived from CRYSTALS-Dilithium. It is constructed over module lattices and uses polynomial arithmetic in the ring $R_q = \mathbb{Z}_q[X]/(X^{256}+1)$ with $q = 8380417$. On constrained processors, the dominant implementation cost comes from repeated Number Theoretic Transform (NTT)-based polynomial operations and modular arithmetic, which have been identified as the main performance bottlenecks in embedded post-quantum implementations across Cortex-M and related platforms~\cite{Benchmark2022,pqm4-benchmarks,Barrett2023,Wei2025}.

For this paper, two ML-DSA properties are the most relevant. First, the standardized signatures are much larger than classical signatures, which directly affects message expansion in lightweight IoT protocols~\cite{NIST-FIPS-204}. Second, signature generation uses rejection sampling and is therefore iterative, which increases latency and introduces runtime variability~\cite{NIST-FIPS-204}. Recent migration and implementation studies further indicate that practical deployment must account not only for raw performance but also for side-channel and fault resilience during the transition from classical signatures to ML-DSA~\cite{Li2024,Dinu2025}. NIST defines three parameter sets with different security and resource trade-offs, summarized in Table~\ref{tab:mldsa-params}.

\begin{table}[h]
	\centering
	\caption{ML-DSA Parameter Sets}
	\label{tab:mldsa-params}
	\begin{tabular}{lccc}
		\toprule
		\textbf{Parameter}  & \textbf{ML-DSA-44} & \textbf{ML-DSA-65} & \textbf{ML-DSA-87} \\
		\midrule
		Security Category   & 2 (AES-128)        & 3 (AES-192)        & 5 (AES-256)        \\
		Matrix $(k,\ell)$   & $(4,4)$            & $(6,5)$            & $(8,7)$            \\
		Private Key (bytes) & 2,560              & 4,032              & 4,896              \\
		Public Key (bytes)  & 1,312              & 1,952              & 2,592              \\
		Signature (bytes)   & 2,420              & 3,309              & 4,627              \\
		\bottomrule
	\end{tabular}
\end{table}

\subsection{MQTT Protocol}

Message Queuing Telemetry Transport (MQTT) is a lightweight publish-subscribe protocol widely used in resource-constrained IoT systems. Its appeal comes from compact control packets, simple broker-mediated communication, and support for multiple Quality of Service levels. However, MQTT does not provide native digital signature fields, so application-layer mechanisms are required when message-level authenticity and non-repudiation are needed~\cite{Samandari2023,Rampazzo2023}. Recent work has also extended post-quantum protection to MQTT-family protocols such as MQTT-SN, explored KEM-oriented MQTT designs for 8-bit sensor nodes, and identified messaging protocols as an active but still sparse part of the broader PQC migration literature~\cite{BlancoRomero2024,Kim2025,Egbuagha2025}.

This design makes MQTT especially sensitive to the deployment cost of ML-DSA. In many telemetry applications, payloads are only a few tens of bytes, whereas ML-DSA signatures occupy several kilobytes. As a result, signature attachment can dominate packet size, increase transmission overhead, and add substantial signing and verification delay on low-power microcontrollers~\cite{Samandari2023,Rampazzo2023,Kim2025,BlancoRomero2024}. This concern is consistent with recent embedded deployment studies beyond MQTT, which likewise report that signature latency, memory footprint, and transmitted bytes remain the main barriers to post-quantum adoption on constrained nodes~\cite{Banegas2021,Marchsreiter2024,pqm4-benchmarks}.

These two background elements define the main evaluation dimensions of this paper. The computational structure of ML-DSA determines signing and verification cost on Cortex-M4 microcontrollers, while the large signature size directly affects MQTT packet expansion, transmission delay, and buffer pressure. Therefore, practical deployment must be assessed jointly at the cryptographic and protocol levels rather than from algorithm specifications alone~\cite{Khalid2019,Ghosh2019,pqm4-benchmarks}.

\section{Optimized Implementation and MQTT-Oriented Integration}\label{sec:architecture}

This section describes the implementation method used to realize the first two contributions stated in Section~I. The emphasis is placed on the algorithmic structure of the ML-DSA software, the implementation choices used to reduce the cost of the dominant arithmetic kernels, and the application-layer framing by which the resulting signatures can be carried in MQTT communication.

\subsection{Algorithmic Overview}

For a parameter set with matrix dimensions $(k,\ell)$, the implementation follows the standard ML-DSA flow of key generation, signing, and verification over the ring $R_q = \mathbb{Z}_q[X]/(X^{256}+1)$. From an initial seed $\xi$, the internal seeds $(\rho,\rho',K)$ are derived and used to generate the public matrix $\mathbf{A}$ and the short secret vectors $\mathbf{s}_1$ and $\mathbf{s}_2$. In transform notation, the key-generation stage is summarized as

\begin{align}
\hat{\mathbf{A}} &= \mathsf{ExpandA}(\rho), \\
\mathbf{s}_1 &= \mathsf{ExpandS}(\rho',\eta,0), \\
\mathbf{s}_2 &= \mathsf{ExpandS}(\rho',\eta,\ell).
\end{align}

\begin{equation}
\mathbf{t} = \operatorname{NTT}^{-1}\!\left(\hat{\mathbf{A}}\hat{\mathbf{s}}_1\right) + \mathbf{s}_2,
\qquad
\hat{\mathbf{s}}_1 = \operatorname{NTT}(\mathbf{s}_1),
\end{equation}

followed by $\mathsf{Power2Round}(\mathbf{t}) = (\mathbf{t}_1,\mathbf{t}_0)$. This decomposition separates the public component $\mathbf{t}_1$ from the low-order private component $\mathbf{t}_0$ and prepares the state used in subsequent signing and verification.

To reduce repeated preprocessing, the implementation keeps the long-lived transformed state

\begin{equation}
\hat{\mathbf{A}}, \quad \hat{\mathbf{s}}_1, \quad \hat{\mathbf{s}}_2, \quad \hat{\mathbf{t}}_0, \quad \widehat{2^d\mathbf{t}_1}
\end{equation}

available for reuse across signature generation and verification. This choice is especially important because signing is iterative and repeatedly invokes the same matrix--vector and challenge-related polynomial products.

\subsection{Implementation-Level Optimization}

The implementation-level optimization is applied without changing the ML-DSA algorithm or parameter sets. As discussed in Section~\ref{sec:complexity}, the dominant cost arises from repeated transforms, modular multiplication, coefficient decomposition, and the rejection-sampling loop in signing. The optimization therefore targets these paths directly while preserving the standardized procedure in FIPS 204~\cite{NIST-FIPS-204}.

The first optimization direction is specialization of the Number Theoretic Transform (NTT) path. Since the polynomial degree is fixed at $n=256$, the transform always consists of eight radix-2 butterfly layers. The basic butterfly update can be written as

\begin{equation}
(u,v) \mapsto (u + \zeta_i v,\; u - \zeta_i v) \Mod{q},
\label{eq:ntt-butterfly}
\end{equation}

where $\zeta_i$ denotes the corresponding twiddle factor. By using a fixed transform schedule and precomputing the required twiddle sequence, the transform avoids unnecessary runtime control overhead. This organization is important because polynomial multiplication is reduced to

\begin{equation}
\widehat{f g} = \hat{f} \odot \hat{g},
\label{eq:pointwise-mul}
\end{equation}

where $\odot$ denotes coefficient-wise multiplication in the NTT domain. Equation~\eqref{eq:pointwise-mul} replaces quadratic-time schoolbook multiplication with the standard NTT-based evaluation path.

The second optimization direction is removal of division from the main arithmetic path. Modular reduction is organized around Barrett-style reciprocal multiplication,

\begin{equation}
\operatorname{red}_q(x) = x - \left\lfloor \frac{x\mu_q}{2^{\kappa}} \right\rfloor q,
\qquad
\mu_q = \left\lfloor \frac{2^{\kappa}}{q} \right\rfloor,
\label{eq:barrett}
\end{equation}

which replaces hardware division with multiplication, bit shifting, and a final correction step. A related reciprocal-based procedure is used when the decomposition stage requires quotient extraction with respect to $2\gamma_2$. This choice is relevant because $\mathsf{Decompose}$, $\mathsf{HighBits}$, and $\mathsf{LowBits}$ are called repeatedly in both signing and verification.

The third optimization direction is reuse of transformed state across iterations of the signing loop. For each counter value $\kappa$, only the fresh masking vector $\mathbf{y}$ and challenge polynomial $c$ are transformed online. The main per-iteration computations are

\begin{align}
\mathbf{y} &= \mathsf{ExpandMask}(\rho'',\kappa), \\
\mathbf{w} &= \operatorname{NTT}^{-1}\!\left(\hat{\mathbf{A}}\hat{\mathbf{y}}\right), \\
\mathbf{z} &= \mathbf{y} + \operatorname{NTT}^{-1}\!\left(\hat{c}\,\hat{\mathbf{s}}_1\right), \\
\mathbf{r}_0 &= \mathsf{LowBits}\!\left(\mathbf{w} - \operatorname{NTT}^{-1}\!\left(\hat{c}\,\hat{\mathbf{s}}_2\right)\right),
\label{eq:sign-core}
\end{align}

where $\hat{\mathbf{y}} = \operatorname{NTT}(\mathbf{y})$ and $\hat{c} = \operatorname{NTT}(c)$. Once the norm constraints for $\mathbf{z}$ and $\mathbf{r}_0$ are satisfied, the hint-generation step is computed as

\begin{align}
\mathbf{u} &= \operatorname{NTT}^{-1}\!\left(\hat{c}\,\hat{\mathbf{t}}_0\right), \\
\mathbf{h} &= \mathsf{MakeHint}\!\left(-\mathbf{u},\; \mathbf{w} - \operatorname{NTT}^{-1}\!\left(\hat{c}\,\hat{\mathbf{s}}_2\right) + \mathbf{u}\right).
\label{eq:hint-step}
\end{align}

Because $\hat{\mathbf{A}}$, $\hat{\mathbf{s}}_1$, $\hat{\mathbf{s}}_2$, and $\hat{\mathbf{t}}_0$ are reused, the rejection-sampling loop does not repeatedly transform long-lived key-dependent state. This optimization directly reduces the overhead of the most expensive operation in ML-DSA deployment.

Algorithm~\ref{alg:opt-sign} summarizes the optimized signing flow. The main implementation feature is that only the ephemeral values generated within the current signing attempt are transformed online, while the matrix and key-dependent NTT-domain state are reused across iterations.

\begin{algorithm}[t]
\caption{Cached-Transform ML-DSA Signing Flow}
\label{alg:opt-sign}
\small
\KwIn{$\mu$, $\rho''$, $\hat{\mathbf{A}}$, $\hat{\mathbf{s}}_1$, $\hat{\mathbf{s}}_2$, $\hat{\mathbf{t}}_0$}
\KwOut{$\sigma=(\tilde{c},\mathbf{z},\mathbf{h})$}
\For{$\kappa \gets 0,\ell,2\ell,\ldots$}{
    $\mathbf{y} \gets \mathsf{ExpandMask}(\rho'',\kappa)$; $\hat{\mathbf{y}} \gets \operatorname{NTT}(\mathbf{y})$\;
    $\mathbf{w} \gets \operatorname{NTT}^{-1}(\hat{\mathbf{A}}\hat{\mathbf{y}})$; $\mathbf{w}_1 \gets \mathsf{HighBits}(\mathbf{w})$\;
    $\tilde{c} \gets H(\mu \parallel \mathsf{Encode}(\mathbf{w}_1))$; $c \gets \mathsf{SampleInBall}(\tilde{c})$\;
    $\hat{c} \gets \operatorname{NTT}(c)$\;
    $\mathbf{a} \gets \operatorname{NTT}^{-1}(\hat{c}\,\hat{\mathbf{s}}_1)$; $\mathbf{b} \gets \operatorname{NTT}^{-1}(\hat{c}\,\hat{\mathbf{s}}_2)$\;
    $\mathbf{z} \gets \mathbf{y} + \mathbf{a}$; $\mathbf{r}_0 \gets \mathsf{LowBits}(\mathbf{w}-\mathbf{b})$\;
    \If{$\|\mathbf{z}\|_{\infty} \geq \gamma_1-\beta$ \textbf{or} $\|\mathbf{r}_0\|_{\infty} \geq \gamma_2-\beta$}{
        continue\;
    }
    $\mathbf{u} \gets \operatorname{NTT}^{-1}(\hat{c}\,\hat{\mathbf{t}}_0)$\;
    $\mathbf{h} \gets \mathsf{MakeHint}(-\mathbf{u},\mathbf{w}-\mathbf{b}+\mathbf{u})$\;
    \If{$\|\mathbf{u}\|_{\infty} \geq \gamma_2$ \textbf{or} $\|\mathbf{h}\|_1 > \omega$}{
        continue\;
    }
    \Return{$(\tilde{c},\mathbf{z},\mathbf{h})$}\;
}
\end{algorithm}

The fourth optimization direction is compact fixed-length serialization. For each parameter set, public keys, private keys, and signatures are encoded as contiguous byte strings with predetermined lengths. In particular,

\begin{align}
|\sigma| &\in \{2420,\ 3309,\ 4627\}\ \text{bytes}, \\
|pk| &\in \{1312,\ 1952,\ 2592\}\ \text{bytes},
\label{eq:enc-len}
\end{align}

for ML-DSA-44, ML-DSA-65, and ML-DSA-87, respectively. This property is useful for constrained implementations because it simplifies buffer management and provides a direct interface to transport protocols that treat the signature as opaque application data.

\subsection{MQTT-Oriented Signature Integration}

MQTT does not define native fields for application-layer digital signatures, so signature transport is handled by payload encapsulation rather than protocol modification. Let $m$ denote the application payload and let $\sigma$ denote the encoded ML-DSA signature. A broker-compatible signed message can be represented as

\begin{align}
\mathcal{F} ={}& \mathsf{type} \parallel \mathsf{len}(m) \parallel m \parallel \mathsf{alg} \parallel \\
& \mathsf{kid} \parallel \mathsf{ts} \parallel \mathsf{len}(\sigma) \parallel \sigma.
\label{eq:mqtt-frame}
\end{align}

where $\mathsf{alg}$ is the algorithm identifier, $\mathsf{kid}$ is the key identifier, and $\mathsf{ts}$ is a freshness field such as a timestamp or sequence value. The signed content can be bound to these metadata fields by defining the authenticated byte string as

\begin{equation}
m^{\star} = m \parallel \mathsf{alg} \parallel \mathsf{kid} \parallel \mathsf{ts}.
\label{eq:auth-bytes}
\end{equation}

At the publisher, the endpoint computes $\sigma = \mathsf{Sign}(sk,m^{\star})$ and places the resulting byte string into the MQTT payload. At the subscriber, the same frame is parsed, freshness and key selection are checked, and verification is performed by evaluating $\mathsf{Verify}(pk,m^{\star},\sigma)$. Because the broker processes the payload as opaque data, the publish-subscribe semantics, topic handling, and broker forwarding logic remain unchanged.

This integration design is intentionally MQTT-oriented rather than MQTT-modifying. The main protocol effect is therefore limited to payload expansion and endpoint-side cryptographic delay, which can be analyzed independently of any broker implementation changes.


\section{Experimental Evaluation}\label{sec:evaluation}

\subsection{Methodology}\label{sec:methodology}

\subsubsection{Measurement Framework}

Performance profiling uses the ARM Cortex-M4 Data Watchpoint and Trace (DWT) unit for cycle-accurate measurement through the \texttt{DWT\_CYCCNT} register. This hardware-based approach avoids software profiling overhead and provides single-cycle temporal resolution. Measurement accuracy was validated by comparison with external logic-analyzer traces, confirming $\pm$1 cycle precision.

Memory-utilization analysis combines static and dynamic measurement techniques. Static memory consumption is quantified with the \texttt{arm-none-eabi-size} toolchain utilities. Dynamic memory profiling uses stack watermarking with 0xDEADBEEF sentinel values at 32-byte intervals. Energy consumption is assessed with INA219 current sensor modules that measure supply current at a sampling frequency of 100\,Hz.

\subsubsection{Software Configuration}

The software environment uses ARM GCC toolchain version 10.3.1 with \texttt{-O3} optimization. Preliminary testing showed that \texttt{-O3} improved performance by 18--23\% over \texttt{-O2}, at the cost of a 12--15\% increase in code size. ML-DSA implementations use the pqm4 reference library with verified NIST test vector compliance. ECDSA baseline measurements use the micro-ecc library with NIST P-256 curves.

MQTT protocol integration uses the Eclipse Paho MQTT Embedded C client communicating with a Mosquitto MQTT broker (version 2.0.15). The network infrastructure uses IEEE 802.11n in the 2.4-GHz band with controlled signal strength from -45 to -52 dBm and packet loss below 0.1\%.

\subsubsection{Benchmark Design}

Each parameter set is evaluated for representative IoT message payloads of 10 bytes (single-sensor readings, 23\% of observed traffic), 50 bytes (multi-parameter telemetry, 51\%), and 100 bytes (diagnostic reports, 18\%). Statistical reliability is supported by 1,000 repeated measurements with interquartile-range-based outlier elimination, which yields rejection rates of 0.8--2.3\%. All measurements are conducted under temperature-controlled conditions (25$^\circ$C $\pm$2$^\circ$C) with device voltage stabilized at 3.3 V $\pm$1\%.

\subsection{Results and Analysis}\label{sec:results}

\subsubsection{Computational Performance}

Computational performance measurements use the pqm4 library implementation with the optimization techniques described in Section~\ref{sec:architecture}. All reported cycle counts correspond to optimized implementations that provide a 40--50\% performance improvement over the reference implementations.

\subsubsection{Key Generation}

Table~\ref{tab:keygen-performance} presents key generation performance across ML-DSA parameter sets.

\begin{table}[h]
	\centering
	\caption{Key Generation Performance (168 MHz)}
	\label{tab:keygen-performance}
	\begin{tabular}{lcccc}
		\toprule
		\textbf{Scheme} & \textbf{Cycles} & \textbf{Time} & \textbf{Ops/s} & \textbf{Overhead} \\
		\midrule
		ECDSA P-256     & 252,000         & 1.50\,ms      & 666.7          & 1.00$\times$      \\
		ML-DSA-44       & 25,368,000      & 151\,ms       & 6.6            & 100.7$\times$     \\
		ML-DSA-65       & 41,832,000      & 249\,ms       & 4.0            & 166.0$\times$     \\
		ML-DSA-87       & 59,976,000      & 357\,ms       & 2.8            & 238.0$\times$     \\
		\bottomrule
	\end{tabular}
\end{table}

The measured 100.7--238$\times$ overhead reflects the computational cost of lattice-based cryptography. Execution times of 151--357\,ms indicate that key generation is feasible for infrequent operations such as device provisioning, but not for high-frequency key-rotation strategies.

\subsubsection{Signature Generation}

Signature generation is the primary performance bottleneck. Table~\ref{tab:sign-performance} reports the signing results.

\begin{table}[h]
	\centering
	\caption{Signature Generation Performance (168 MHz)}
	\label{tab:sign-performance}
	\begin{tabular}{lcccc}
		\toprule
		\textbf{Scheme} & \textbf{Payload} & \textbf{Time} & \textbf{Ops/s} & \textbf{Overhead} \\
		\midrule
		ECDSA P-256     & 10\,B            & 9.19\,ms      & 108.8          & 1.00$\times$      \\
		ECDSA P-256     & 50\,B            & 9.39\,ms      & 106.5          & 1.00$\times$      \\
		ECDSA P-256     & 100\,B           & 9.59\,ms      & 104.3          & 1.00$\times$      \\
		\midrule
		ML-DSA-44       & 50\,B            & 663\,ms       & 1.51           & 70.6$\times$      \\
		ML-DSA-65       & 50\,B            & 853\,ms       & 1.17           & 90.8$\times$      \\
		ML-DSA-87       & 50\,B            & 1,136\,ms     & 0.88           & 121.0$\times$     \\
		\bottomrule
	\end{tabular}
\end{table}

Signing latencies of 663--1,136\,ms across parameter sets remain 70.6--121.0$\times$ higher than those of ECDSA despite optimization. Variability caused by rejection sampling introduces timing non-determinism and therefore requires worst-case latency analysis in real-time applications.

\subsubsection{Signature Verification}

Table~\ref{tab:verify-performance} presents verification latency measurements.

\begin{table}[h]
	\centering
	\caption{Signature Verification Performance (168 MHz)}
	\label{tab:verify-performance}
	\begin{tabular}{lcccc}
		\toprule
		\textbf{Scheme} & \textbf{Payload} & \textbf{Time} & \textbf{Ops/s} & \textbf{Overhead} \\
		\midrule
		ECDSA P-256     & 50\,B            & 16.2\,ms      & 61.7           & 1.00$\times$      \\
		ML-DSA-44       & 50\,B            & 420\,ms       & 2.38           & 25.9$\times$      \\
		ML-DSA-65       & 50\,B            & 533\,ms       & 1.88           & 32.9$\times$      \\
		ML-DSA-87       & 50\,B            & 710\,ms       & 1.41           & 43.8$\times$      \\
		\bottomrule
	\end{tabular}
\end{table}

Verification executes deterministically because it does not involve rejection sampling, which yields more predictable latency characteristics. Verification incurs 26--44$\times$ overhead relative to ECDSA, compared with 70--122$\times$ for signature generation.

\subsubsection{Memory Utilization}

Table~\ref{tab:memory} presents static and dynamic memory requirements.

\begin{table}[h]
	\centering
	\caption{Memory Requirements}
	\label{tab:memory}
	\begin{tabular}{lcccc}
		\toprule
		\textbf{Scheme} & \textbf{Flash} & \textbf{Stack} & \textbf{Keys} & \textbf{Total SRAM} \\
		\midrule
		ECDSA P-256     & 9.7\,KB        & 0.8\,KB        & 0.1\,KB       & 2.1\,KB             \\
		ML-DSA-44       & 37.2\,KB       & 6.4\,KB        & 3.8\,KB       & 22.7\,KB            \\
		ML-DSA-65       & 54.8\,KB       & 8.7\,KB        & 5.8\,KB       & 32.8\,KB            \\
		ML-DSA-87       & 73.9\,KB       & 11.2\,KB       & 7.3\,KB       & 43.1\,KB            \\
		\bottomrule
	\end{tabular}
\end{table}

Memory requirements of 22.7--43.1\,KB SRAM constrain deployment to mid-range microcontrollers. Flash requirements of 37.2--73.9\,KB are still compatible with some constrained platforms, but they reduce flexibility in cost-sensitive IoT deployments.

\subsubsection{Protocol-Level Overhead}

\paragraph{Message Size}

Table~\ref{tab:message-overhead} quantifies message size overhead for signed MQTT payloads.

\begin{table}[h]
	\centering
	\caption{MQTT Message Size Overhead}
	\label{tab:message-overhead}
	\begin{tabular}{lcccc}
		\toprule
		\textbf{Scheme} & \textbf{Payload} & \textbf{Signed} & \textbf{Unsigned} & \textbf{Ratio} \\
		\midrule
		ECDSA P-256     & 50\,B            & 122\,B          & 58\,B             & 2.1$\times$    \\
		ML-DSA-44       & 50\,B            & 2,478\,B        & 58\,B             & 42.7$\times$   \\
		ML-DSA-65       & 50\,B            & 3,367\,B        & 58\,B             & 58.1$\times$   \\
		ML-DSA-87       & 50\,B            & 4,685\,B        & 58\,B             & 80.8$\times$   \\
		\bottomrule
	\end{tabular}
\end{table}

This message-size overhead directly affects network bandwidth consumption and transmission cost in cellular IoT deployments where charging is based on transferred bytes.

\paragraph{End-to-End Latency}

Table~\ref{tab:e2e-latency} presents complete publish-subscribe workflow latency incorporating signature generation, network transmission, and verification.

\begin{table}[h]
	\centering
	\caption{End-to-End MQTT Latency}
	\label{tab:e2e-latency}
	\begin{tabular}{lccccc}
		\toprule
		\textbf{Scheme} & \textbf{Sign} & \textbf{Net.} & \textbf{Verify} & \textbf{Total} & \textbf{Overhead} \\
		\midrule
		ECDSA P-256     & 9.4\,ms       & 28.5\,ms      & 16.2\,ms        & 54.1\,ms       & 1.00$\times$      \\
		ML-DSA-44       & 663\,ms       & 31.2\,ms      & 420\,ms         & 1,114\,ms      & 20.6$\times$      \\
		ML-DSA-65       & 853\,ms       & 33.8\,ms      & 533\,ms         & 1,420\,ms      & 26.2$\times$      \\
		ML-DSA-87       & 1,136\,ms     & 37.4\,ms      & 710\,ms         & 1,883\,ms      & 34.8$\times$      \\
		\bottomrule
	\end{tabular}
\end{table}

End-to-end latency of 1.11--1.88 s exceeds sub-second bounds for interactive IoT applications and therefore requires architectural accommodations in latency-sensitive deployments.

\paragraph{Sustainable Throughput}

Sustainable message rates range from 0.87 to 1.52 messages/s across ML-DSA parameter sets, compared with 104--109 messages/s for ECDSA. Signature generation is the primary performance bottleneck and causes a throughput reduction of 70--122$\times$ relative to the classical baseline.

\subsubsection{Trade-off Analysis}

Table~\ref{tab:parameter-tradeoffs} quantifies security-performance trade-offs normalized to ML-DSA-44.

\begin{table}[h]
	\centering
	\caption{ML-DSA Parameter Set Trade-offs (Normalized to ML-DSA-44)}
	\label{tab:parameter-tradeoffs}
	\begin{tabular}{lccc}
		\toprule
		\textbf{Metric}   & \textbf{DSA-44} & \textbf{DSA-65} & \textbf{DSA-87} \\
		\midrule
		Security Level    & 2 (AES-128)     & 3 (AES-192)     & 5 (AES-256)     \\
		Signing Time      & 1.00$\times$    & 1.30$\times$    & 1.75$\times$    \\
		Verification Time & 1.00$\times$    & 1.27$\times$    & 1.72$\times$    \\
		Signature Size    & 1.00$\times$    & 1.37$\times$    & 1.91$\times$    \\
		Total SRAM        & 1.00$\times$    & 1.45$\times$    & 1.90$\times$    \\
		\bottomrule
	\end{tabular}
\end{table}

For IoT deployments with 5--10 year operational lifetimes under current quantum-computing development trajectories, NIST security level 2 (ML-DSA-44) provides a reasonable balance between quantum resistance and resource cost. Long-term infrastructure deployments exceeding 20 years and requiring more conservative security margins may justify ML-DSA-87 despite its 75--90\% increase in resource overhead.

\subsubsection{Deployment Feasibility}

Deployment feasibility is evaluated across representative IoT application categories based on the measured performance characteristics.

\textbf{High-throughput sensor networks}: Frequent authenticated message publication creates a computational bottleneck that limits sustainable publication rates to 0.87--1.52 messages/s across ML-DSA parameter sets, compared with 104--109 messages/s for ECDSA. Deployments requiring publication rates above 1 message/s therefore need mitigation strategies such as message aggregation, publisher-side caching of unchanged states, or hybrid authentication in which ML-DSA is reserved for security-critical messages and MAC-based authentication is used for routine telemetry. For example, a network with 100 sensor nodes, each publishing at 0.01 Hz, yields an aggregate throughput of 1 message/s and remains within the computational capacity of ML-DSA-44.

\textbf{Battery-powered devices}: Energy-constrained devices require additional consideration of signing cost. A signature-generation cost of 110.4 million cycles at 168 MHz corresponds to 657 ms for ML-DSA-44 and implies power draw of approximately 50 mW, assuming a 3.3-V supply and 15-mA active current. The resulting energy per signature is 32.9 mJ. For a device that transmits 100 authenticated messages per day from a 2,000-mAh battery (3.3-V nominal voltage, 23.8 kJ total energy), ML-DSA signature generation consumes 3.29 J per day, or 13.8\% of the total energy budget. This estimate excludes network-transmission energy and verification overhead at subscriber devices. ML-DSA therefore remains viable for battery-powered publishers with low daily message frequencies, whereas higher publication rates require energy harvesting or mains power.

\textbf{Real-time control systems}: Systems with sub-second latency requirements violate their timing constraints under ML-DSA end-to-end authentication latency of 1.11--1.88 s. Deployments requiring real-time authenticated command-response interactions therefore need architectural accommodations such as asymmetric publisher-subscriber security models, pre-signed command templates for restricted command spaces, or hybrid cryptographic modes that use classical signatures for time-critical operations and periodic ML-DSA re-authentication for long-term security. Pure ML-DSA authentication remains suitable for monitoring applications that tolerate multi-second latency, but it is not appropriate for closed-loop control systems requiring deterministic sub-second response.

\subsubsection{Optimization Opportunities}

Several optimization strategies can mitigate ML-DSA performance overhead in resource-constrained deployments.

\textbf{Parameter set selection}: Deployment-specific security analysis may support the use of ML-DSA-44, which reduces signing latency by 41.6\% (1,136\,ms $\rightarrow$ 663\,ms), signature size by 47.7\% (4,627 bytes $\rightarrow$ 2,420 bytes), and memory consumption by 47.3\% (43.1\,KB $\rightarrow$ 22.7\,KB) relative to ML-DSA-87 while maintaining NIST security level 2 protection.

\textbf{Message aggregation}: Batch signature generation over aggregated sensor readings amortizes cryptographic overhead across multiple measurements. Aggregating $N$ measurements into a single signed payload reduces the per-measurement signing overhead from 657\,ms to $657/N$\,ms, at the cost of an $N\times$ increase in sampling-interval latency.

\textbf{Hybrid authentication}: Selective ML-DSA deployment for security-critical messages combined with MAC-based authentication for routine telemetry reduces average authentication overhead while preserving non-repudiation for critical operations. For a traffic mix of 90\% routine telemetry (MAC: 0.5\,ms) and 10\% critical messages (ML-DSA-44: 657\,ms), the weighted average overhead is 66.2\,ms, compared with 657\,ms for pure ML-DSA deployment.

\textbf{Hardware acceleration}: Future ARM Cortex-M processors with cryptographic acceleration support for NTT operations could provide 5--10$\times$ performance improvements and reduce ML-DSA signing latency to below 100 ms.

\section{Conclusion}\label{sec:conclusion}

This paper presents a performance characterization of ML-DSA integration in MQTT-based IoT systems on ARM Cortex-M4 microcontrollers. Signature-generation latencies of 663--1,136\,ms correspond to 70.6--121.0$\times$ overhead relative to ECDSA P-256 and identify signing as the primary computational bottleneck, limiting sustainable message throughput to 0.87--1.52 messages/s. Memory requirements of 22.7--43.1\,KB SRAM and 37.2--73.9\,KB Flash further constrain deployment to mid-range microcontrollers.

ML-DSA remains viable for applications that tolerate multi-second latency and sub-hertz publication frequencies, including environmental monitoring, asset tracking, and periodic telemetry reporting. Real-time control systems that require sub-second response need architectural accommodations or hybrid authentication approaches. Future work includes evaluating hardware acceleration on emerging ARM Cortex-M processors with cryptographic extensions and studying batch-signature-verification schemes for gateway devices.

\bibliographystyle{IEEEtran_doi}
\bibliography{ biblio.bib }

\begin{IEEEbiographynophoto}{Jiahao Xiang}
	is pursuing a Master's degree in Electronic Information at Hengyang Normal University, China. His research focuses on cryptographic engineering and efficient implementations of block ciphers on resource-constrained devices. Publications include works on lightweight cryptography optimization and contributions to open-source cryptographic projects.
\end{IEEEbiographynophoto}

\vskip -2\baselineskip plus -1fil

\begin{IEEEbiographynophoto}{Lang Li}
	received his Ph.D. and Master's degrees in computer science from Hunan University, Changsha, China, in 2010 and 2006, respectively, and earned his B.S. degree in circuits and systems from Hunan Normal University in 1996. Since 2011, he has been working as a professor in the College of Computer Science and Technology at the Hengyang Normal University, Hengyang, China. He has research interests in embedded system and information security.
\end{IEEEbiographynophoto}

\vskip -2\baselineskip plus -1fil

\begin{IEEEbiographynophoto} {Jingya Feng}
	received the M.S. degree from Hunan Normal University, Changsha, China, in 2021, and the Ph.D. degree from Guilin University of Electronic Science and Technology, Guilin, China, in 2025. She Joined the School of Computer Science and Technology, Hengyang Normal University in July 2025. Her main research interests include cryptology, with particular focus on the design and optimized implementation of symmetric encryption schemes.
\end{IEEEbiographynophoto}

\end{document}
